---
title: "Data607_Proj2"
author: "Joe_Rovalino, Tamiko Jenkins and Nicholas Chung"
date: "10/02/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(getPass)
library(RMySQL)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
```

## R Markdown

Relevant Information:
The Powerball lottery was loaded into MySQL DB and describes Powerball winning numbers over a period of Feb 3rd to the present. 
(1) Created a MySQL database that includes all of the information above. - CHOSE to load to MySQL and use the lesson from homework 2 to create DB.
(2) Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy
and transform your data.
(3) Perform analysis to compare the numbers in the various position of the powerball winning number.
(4) Code will be in an R Markdown file, posted to rpubs.com, and will include narrative
descriptions of your data cleanup work, analysis, and conclusions. We will include in our
project submission:
The URL to the .Rmd file in your GitHub repository. and
The URL for your rpubs.com web page.

```{r}
db_user <- 'root'
db_password <- getPass::getPass("Enter the password: ") 
db_name <- 'data607proj2'
db_table <- 'powerball'
db_host <- '127.0.0.1' # for local access
db_port <- 3306

mydb <-  dbConnect(MySQL(), user = db_user, password = db_password,
                 dbname = db_name, host = db_host, port = db_port)

s <- paste0("select * from ", db_table)
rs <- dbSendQuery(mydb, s)
df <-  fetch(rs, n = -1)
on.exit(dbDisconnect(mydb))

df
```

# load csv into mem for ad hoc EDA. 
##### NC note: happy to host the mysql db in my aws and share pw via lastpass so we can mutually reference a common table.
```{r}
raw_data <- read.csv("powerball.csv", sep = ",", header = FALSE)
nrow(na.omit(raw_data))
head(raw_data)
```

# initial observations on state of raw data:
* we know that there is variance in header count, where data from 2012 to 2014 has one less header "Power Play"
* rows that represent drawings from 2012 to 2014 have value "0" where header is "Power Play" / column position is 11
* column 1 is not useful

# scrape table with estimated jackpot winnings
```{r}
library(xml2)
library(rvest)
url <- "https://www.txlottery.org/export/sites/lottery/Games/Powerball/Winning_Numbers/print.html_2013354932.html"
winnings <- xml2::read_html(url)
winnings <- rvest::html_table(winnings)[[1]]
nrow(na.omit(raw_data))
head(winnings)
```

# observations on both data sets collected so far:
* after omitting rows with missing data, there is row count parity
* "estimated jackpot" column contains values that require transformation into numeric representation
* "jackpot option" column contains "CVO" value, which indicates whether the winner took the lump sum over annual payments
* "winning numbers" column headers will require transformation in unique values by position
* in "jackpot winners" column, value "roll" indicates a roll-over, where no winner was selected (ref: https://answers.yahoo.com/question/index?qid=20110429210210AA8Y824&guccounter=1&guce_referrer=aHR0cHM6Ly93d3cuZWNvc2lhLm9yZy8&guce_referrer_sig=AQAAAB6jQ-NGhdyMOMlFbNTSrqAgIkL-1JtAQoQyaGrwCWYspY8f3mDVzHJjsZ0enRMo17Ww9cYAzc2fmAKhcEkhTGArfXiiOZfMk1Du9658VBZZg5ZU8MjAsbaLY15qkJ4zaGU10aXtW5Fqm0qsgJvVSsPIX5_CI5veEM03FoI_h1lC). 
* where "jackpot option" value is "CVO", it is also possible to spot a value indicating the number of "jackpot winners"

# update headers 
```{r}
names(winnings) <- c("draw_date", "num1", "num2", "num3", "num4", "num5", "powerball", "power_play", "est_jackpot", "jackpot_winners", "jackpot_options")
winnings
```

# transform "est_jackpot" into numeric values
```{r}
table(winnings$est_jackpot) # check distribution of values to identify values that need replacing
winnings$est_jackpot <- sub(" Million", "000000", winnings$est_jackpot) # replace "Million" and preceding space with appropriate digits
winnings$est_jackpot <- sub(" Billion", "000000000", winnings$est_jackpot) # replace "Million" and preceding space with appropriate digits
winnings$est_jackpot <- as.numeric(sub("\\$", "", winnings$est_jackpot)) # transform class to numeric
head(winnings)
winnings
```

# drop rows where winning numbers have alpha to remove descriptive text
```{r}
clean.winnings = filter(winnings, !grepl("First", num1)) # remove rows with string "First", which are unique to descriptive text
head(clean.winnings)
```

# we want "jackpot winners" and "jackpot_options" tidied such that: 
* "outcome": roll, out of state winner, CVO
* "winner_count": number of winners
```{r}
# cast win, type of winner, and number of winners
casted.winnings <- clean.winnings %>%
  pivot_wider(names_from = "jackpot_winners", values_from = "jackpot_winners")
casted.winnings
```

```{r}
# rename all headers for easy reference
names(casted.winnings) <- c("draw_date", "num1", "num2", "num3", "num4", "num5", "powerball", "power_play", "est_jackpot", "jackpot_options", "roll", "out of state winner", "winner_count")

# recombine roll/out of state into "status"
casted.winnings$status <- paste(casted.winnings$roll, casted.winnings$'out of state winner', sep=" ")

# clean up NA values
casted.winnings$status <- sub(" NA", "", casted.winnings$status)
casted.winnings$status <- sub("NA ", "", casted.winnings$status)
casted.winnings$status <- sub("NA", "Winner", casted.winnings$status)

# select only those columns we want to keep
tidy.winnings <- subset(casted.winnings, select=c("draw_date", "num1", "num2", "num3", "num4", "num5", "powerball", "power_play", "est_jackpot", "jackpot_options", "status", "winner_count"))

# replace NA with 0
tidy.winnings$winner_count[is.na(tidy.winnings$winner_count)] <- 0

head(tidy.winnings)
```

```{r}
# transform relevant datatypes to integer and date 
tw <- tidy.winnings %>% mutate(
  draw_date = as.Date(tidy.winnings$draw_date,format="%m/%d/%Y")
  ) %>% 
  mutate_at(
    vars(num1, num2, num3, num4, num5, powerball, power_play, est_jackpot, winner_count), funs(as.integer)
    )
head(tw)
```

```{r}
pb_df <- as_tibble(tw)
pb_df
```

# commence analysis of data
#### questions:
* distribution of numbers (how "random"/normally distributed is number selection?)
* what is distribution of jackpot outcomes? 
* how has growth of jackpot size between wins changed over time?
```{r}
# group the data by position - to detemine number value and count occurance of number in the position
jkpt_temp <- pb_df %>% group_by (status) %>% summarize (n()) 
# rename headers
names(jkpt_temp) <- c("status", "count")
head(jkpt_temp)

ggplot(jkpt_temp, aes(fill=status, y=count, x=status)) +
  ggtitle("jackpot winners") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_bar(position='dodge', stat="identity") +
  xlab('jackpot winners') +
  ylab('count')
```

```{r}
# frequency position integers
freq_num1 <- pb_df %>% group_by (num1) %>% summarize (n())
freq_num2 <- pb_df %>% group_by (num2) %>% summarize (n())
freq_num3 <- pb_df %>% group_by (num3) %>% summarize (n())
freq_num4 <- pb_df %>% group_by (num4) %>% summarize (n())
freq_num5 <- pb_df %>% group_by (num5) %>% summarize (n())
freq_powerball <- pb_df %>% group_by (powerball) %>% summarize (n())

# rename headers
names(freq_powerball) <- c("powerball", "count")
head(freq_powerball)

ggplot(freq_powerball, aes(fill=powerball, y=count, x=powerball)) +
  ggtitle("distribution of drawn powerball numbers") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_bar(position='dodge', stat="identity") +
  xlab('winning numbers') +
  ylab('count')
```

```{r}
ggplot(data=pb_df, aes(x=draw_date, y=est_jackpot, group=1)) +
  ggtitle("jackpot size over time") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line(color="red") + 
  xlab('time') +
  ylab('jackpot size, in $')
```

# findings
* distribution of winning numbers suggests non-random selection