---
title: "Data 607 - Project 2"
author: "Nicholas Chung, Joe Rovalino, T. Jenkins"
subtitle: 'Dataset 2: Immigration data'
output:
  pdf_document:
    toc: yes
  html_document:
    code_download: yes
    code_folding: hide
    highlight: pygments
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(stringr)
```


Relevant Information:
Suggestion of data by classmate - Mengqin Cai 
These data set published by the United Nation, which calculates Female migrants as a percentage of the international migrant stock . The dataset has seven variables,  Area, Country, Notes, Country code, Type of data (a), Years, Percentage. This table compares different countries' female migrants as a percentage of the international migrant stock and shows their changes from 1990 to 2015. The data combine the area and country with the same column and set year values as header names. It is not easy to process the data if we want to compare the data by country and find out the year trend from 1995 to 2015. We call this type of data untidy and need to transform for easy data process.
The URL to the .Rmd file in your GitHub repository. and
The URL for your rpubs.com web page.



## Data 

## load csv into mem for ad hoc EDA. 
## this write up was very helpful - they used a very similiar dataset
## https://blog.exploratory.io/tidy-untidy-united-nations-migration-data-with-tidyr-167cbd24c5c2

```{r}
raw_data <- read.csv("Immigration Data.csv", sep = ",", , skip = 16, header = FALSE) #initial tidy on removing first 16 rows of non #counmtry dat
#?read.csv
head(raw_data)
# V5 column, if blank this is region’ rows don’t have any value ther. We will use this to remove the regions and leave the country data for analysis
df <- raw_data
df
```


# Tidy
```{r}
#gets the region (non-country) info to inspect into a data frame for removal so that we can be left with country data only. 
?filter
 df_not_na <- df %>% filter (V5 != "" )
 df_not_na
# df_not_na is now filtered to the country only. Use names to provide beter column names

names(df_not_na) <- c("Sort_order", "Destination_Country", "Notes", "Country_code", "TypeofData", "a1990_perc", "a1995_perc","a2000_perc","a2005_perc","a2010_perc","a2015_perc","a2017_perc" )
df_not_na
```


# Ordered by Destination Country
``` {r}
arrange(df_not_na, Destination_Country)

df_2017 <- df_not_na %>% select(-a1990_perc, -a1995_perc, -a2000_perc,-a2005_perc,-a2010_perc ,-a2015_perc)
df_2017

df_2017_top5 <- top_n(df_2017,5,a2017_perc)
df_2017_top5


```
