---
title: "Data 607 - Project 2 Draft"
author: "Nicholas Chung, Joe Rovalino, T. Jenkins"
subtitle: 'Dataset 1: Powerball data'
output:
  html_document:
    code_download: yes
    code_folding: hide
    highlight: pygments
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

The goal of this assignment is to give you practice in preparing different datasets for downstream
analysis work.

Your task is to:

(1) Choose any three of the “wide” datasets identified in the Week 6 Discussion items. (You may
use your own dataset; please don’t use my Sample Post dataset, since that was used in your
Week 6 assignment!) 

Relevant Information:

The Powerball lottery was loaded into MySQL DB and describes Powerball winning numbers over a period of Feb 3rd to the present. 
(1) Created a MySQL database that includes all of the information above. - CHOSE to load to MySQL and use the lesson from homework 2 to create DB.
(2) Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy
and transform your data.
(3) Perform analysis to compare the numbers in the various position of the powerball winning number.
(4) Code will be in an R Markdown file, posted to rpubs.com, and will include narrative
descriptions of your data cleanup work, analysis, and conclusions. We will include in our
project submission:
* The URL to the .Rmd file in your GitHub repository. and
* The URL for your rpubs.com web page.

Submission:
Tamiko Jenkins 
Tidying Sequences and Classifications Attachment
Discussion 5: Untidy Data
Find (or create) an example of an untidy dataset—for example, a “wide” table of descriptive data taken from an HTML page.  You should also describe some kind of analysis that might be performed on the data.  You should not tidy, transform, or analyze the data here, since that will be one part of your Project 2 deliverable (which is also available in your week 5 folder!).  Please see my sample post, which is taken from your Week 5 assignment.

The wide dataset (linked and attached) I will work with is from the Texas Lottery site (structure below). Will it is not wide in comparison to many other sets out there, it does illustrate a the concept of gathering columns (Num):

"All Powerball drawings held since Texas began participating in game drawings on February 3, 2010.

The information in the file is organized in the following manner, one row for each drawing:

Feb 3, 2010 to Jan 14, 2012:
Game Name, Month, Day, Year, Num1, Num2, Num3, Num4, Num5, Powerball, Power Play

Jan 18, 2012 to Jan 18, 2014:
Game Name, Month, Day, Year, Num1, Num2, Num3, Num4, Num5, Powerball

Jan 22, 2014 to present:
Game Name, Month, Day, Year, Num1, Num2, Num3, Num4, Num5, Powerball, Power Play"

https://www.txlottery.org/export/sites/lottery/Games/Powerball/Winning_Numbers/download.html

Commonly calculated statistics include frequency and days ago. We can also calculate proportion in winning numbers, associated winnings, common pairs, common triplets, common sequences, odds (based on jackpot), draw day. Lotto statistics are calculated mostly with the goal of winning, but they can also be calculated to detect cheating (by checking for randomness). 

## Data {.tabset}

Initial observations on state of raw data:
* we know that there is variance in header count, where data from 2012 to 2014 has one less header "Power Play"
* rows that represent drawings from 2012 to 2014 have value "0" where header is "Power Play" / column position is 11
* column 1 is not useful

Observations on both data sets collected so far:
* after omitting rows with missing data, there is row count parity
* "estimated jackpot" column contains values that require transformation into numeric representation
* "jackpot option" column contains "CVO" value, which indicates whether the winner took the lump sum over annual payments
* "winning numbers" column headers will require transformation in unique values by position
* in "jackpot winners" column, value "roll" indicates a roll-over, where no winner was selected (ref: https://answers.yahoo.com/question/index?qid=20110429210210AA8Y824&guccounter=1&guce_referrer=aHR0cHM6Ly93d3cuZWNvc2lhLm9yZy8&guce_referrer_sig=AQAAAB6jQ-NGhdyMOMlFbNTSrqAgIkL-1JtAQoQyaGrwCWYspY8f3mDVzHJjsZ0enRMo17Ww9cYAzc2fmAKhcEkhTGArfXiiOZfMk1Du9658VBZZg5ZU8MjAsbaLY15qkJ4zaGU10aXtW5Fqm0qsgJvVSsPIX5_CI5veEM03FoI_h1lC). 
* where "jackpot option" value is "CVO", it is also possible to spot a value indicating the number of "jackpot winners"


### MySQL


db_user <- 'root'
db_password <- getPass::getPass("Enter the password: ") 
db_name <- 'data607proj2'
db_table <- 'powerball'
db_host <- '127.0.0.1' # for local access
db_port <- 3306

mydb <-  dbConnect(MySQL(), user = db_user, password = db_password,
                 dbname = db_name, host = db_host, port = db_port)

s <- paste0("select * from ", db_table)
rs <- dbSendQuery(mydb, s)
df <-  fetch(rs, n = -1)
on.exit(dbDisconnect(mydb))

df



#### Reference
Texas Lottery. Download. Txlottery.org. https://www.txlottery.org/export/sites/lottery/Games/Powerball/Winning_Numbers/download.html. Published 2019. Accessed October 7, 2019.

### CSV

```{r}
raw_data <- read.csv("powerball.csv", sep = ",", header = FALSE)
nrow(na.omit(raw_data))
head(raw_data)
```

#### Reference
Texas Lottery. Download. Txlottery.org. https://www.txlottery.org/export/sites/lottery/Games/Powerball/Winning_Numbers/download.html. Published 2019. Accessed October 7, 2019.

### Scrapped Data

#### Scrape table with estimated jackpot winnings
```{r}
library(xml2)
library(rvest)
url <- "https://www.txlottery.org/export/sites/lottery/Games/Powerball/Winning_Numbers/print.html_2013354932.html"
winnings <- xml2::read_html(url)
winnings <- rvest::html_table(winnings)[[1]]
nrow(na.omit(raw_data))
head(winnings)
```

#### Reference
Powerball Winning Numbers - Print Friendly Format. Txlottery.org. https://www.txlottery.org/export/sites/lottery/Games/Powerball/Winning_Numbers/print.html_2013354932.html. Published 2019. Accessed October 7, 2019.


## Clean and Tidying {.tabset}

### Cleaning

#### update headers 
```{r}
library(tidyr)
library(dplyr)
library(readr)
names(winnings) <- c("draw_date", "num1", "num2", "num3", "num4", "num5", "powerball", "power_play", "est_jackpot", "jackpot_winners", "jackpot_options")
winnings
```

#### transform "est_jackpot" into numeric values
```{r}
table(winnings$est_jackpot) # check distribution of values to identify values that need replacing
winnings$est_jackpot <- sub(" Million", "000000", winnings$est_jackpot) # replace "Million" and preceding space with appropriate digits
winnings$est_jackpot <- sub(" Billion", "000000000", winnings$est_jackpot) # replace "Million" and preceding space with appropriate digits
winnings$est_jackpot <- as.numeric(sub("\\$", "", winnings$est_jackpot)) # transform class to numeric
head(winnings)
winnings
```

#### drop rows where winning numbers have alpha to remove descriptive text
```{r}
clean.winnings = dplyr::filter(winnings, !grepl("First", num1)) # remove rows with string "First", which are unique to descriptive text
head(clean.winnings)
```

### Tidyr

#### We want "jackpot winners" and "jackpot_options" tidied such that: 
* "outcome": roll, out of state winner, CVO
* "winner_count": number of winners

```{r}
# cast win, type of winner, and number of winners
casted.winnings <- clean.winnings %>%
#  pivot_wider(names_from = "jackpot_winners", values_from = "jackpot_winners")  # Requires dev version of tidyr
#  workaround
  gather(names_from = "jackpot_winners", values_from = "jackpot_winners")
library(fastDummies)
dummy_winners <- dummy_cols(casted.winnings$value)
casted.winnings$key <- dummy_winners[,2]
casted.winnings$value <- dummy_winners[,3]
casted.winnings$value2 <- dummy_winners[,3]
head(dummy_winners)
casted.winnings
```

```{r}
# rename all headers for easy reference
names(casted.winnings) <- c("draw_date", "num1", "num2", "num3", "num4", "num5", "powerball", "power_play", "est_jackpot", "jackpot_options", "roll", "out of state winner", "winner_count")

# recombine roll/out of state into "status"
casted.winnings$status <- paste(casted.winnings$roll, casted.winnings$'out of state winner', sep=" ")

# clean up NA values
casted.winnings$status <- sub(" NA", "", casted.winnings$status)
casted.winnings$status <- sub("NA ", "", casted.winnings$status)
casted.winnings$status <- sub("NA", "Winner", casted.winnings$status)

# select only those columns we want to keep
tidy.winnings <- subset(casted.winnings, select=c("draw_date", "num1", "num2", "num3", "num4", "num5", "powerball", "power_play", "est_jackpot", "jackpot_options", "status", "winner_count"))

# replace NA with 0
tidy.winnings$winner_count[is.na(tidy.winnings$winner_count)] <- 0

head(tidy.winnings)
```

```{r}
# transform relevant datatypes to integer and date 
tw <- tidy.winnings %>% mutate(
  draw_date = as.Date(tidy.winnings$draw_date,format="%m/%d/%Y")
  ) %>% 
  mutate_at(
    vars(num1, num2, num3, num4, num5, powerball, power_play, est_jackpot, winner_count), funs(as.integer)
    )
head(tw)
```

```{r}
pb_df <- as_tibble(tw)
pb_df
```

## Analysis  

### Questions:
* Distribution of numbers (how "random"/normally distributed is number selection?)
* What is distribution of jackpot outcomes? 
* How has growth of jackpot size between wins changed over time?

### Group the data by position
Detemine number value and count occurance of number in the position.
```{r}
library(ggplot2)
jkpt_temp <- pb_df %>% group_by (status) %>% summarize (n()) 
# rename headers
names(jkpt_temp) <- c("status", "count")
head(jkpt_temp)

ggplot(jkpt_temp, aes(fill=status, y=count, x=status)) +
  ggtitle("jackpot winners") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_bar(position='dodge', stat="identity") +
  xlab('jackpot winners') +
  ylab('count')
```

### Frequency position integers
```{r}

freq_num1 <- pb_df %>% group_by (num1) %>% summarize (n())
freq_num2 <- pb_df %>% group_by (num2) %>% summarize (n())
freq_num3 <- pb_df %>% group_by (num3) %>% summarize (n())
freq_num4 <- pb_df %>% group_by (num4) %>% summarize (n())
freq_num5 <- pb_df %>% group_by (num5) %>% summarize (n())
freq_powerball <- pb_df %>% group_by (powerball) %>% summarize (n())

# rename headers
names(freq_powerball) <- c("powerball", "count")
head(freq_powerball)
```

### Distribution of drawn powerball numbers
```{r}
ggplot(freq_powerball, aes(fill=powerball, y=count, x=powerball)) +
  ggtitle("distribution of drawn powerball numbers") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_bar(position='dodge', stat="identity") +
  xlab('winning numbers') +
  ylab('count')
```

### Jackpot size over time
```{r}
ggplot(data=pb_df, aes(x=draw_date, y=est_jackpot, group=1)) +
  ggtitle("jackpot size over time") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line(color="red") + 
  xlab('time') +
  ylab('jackpot size, in $')
```







