---
title: "Data607_Proj2"
author: "Joe_Rovalino, Tamiko Jenkins and Nicholas Chung"
date: "10/02/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Relevant Information:
The Powerball lottery was loaded into MySQL DB and describes Powerball winning numbers over a period of Feb 3rd to the present. 
(1) Created a MySQL database that includes all of the information above. - CHOSE to load to MySQL and use the lesson from homework 2 to create DB.
(2) Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy
and transform your data.
(3) Perform analysis to compare the numbers in the various position of the powerball winning number.
(4) Code will be in an R Markdown file, posted to rpubs.com, and will include narrative
descriptions of your data cleanup work, analysis, and conclusions. We will include in our
project submission:
The URL to the .Rmd file in your GitHub repository. and
The URL for your rpubs.com web page.

```{r}
library(getPass)
library(RMySQL)

db_user <- 'root'
db_password <- getPass::getPass("Enter the password: ") 
db_name <- 'data607proj2'
db_table <- 'powerball'
db_host <- '127.0.0.1' # for local access
db_port <- 3306

mydb <-  dbConnect(MySQL(), user = db_user, password = db_password,
                 dbname = db_name, host = db_host, port = db_port)

s <- paste0("select * from ", db_table)
rs <- dbSendQuery(mydb, s)
df <-  fetch(rs, n = -1)
on.exit(dbDisconnect(mydb))

df
```

# load csv into mem for ad hoc EDA. 
##### NC note: happy to host the mysql db in my aws and share pw via lastpass so we can mutually reference a common table.
```{r}
raw_data <- read.csv("powerball.csv", sep = ",", header = FALSE)
raw_data
```

# initial observations on state of raw data:
* we know that there is variance in header count, where data from 2012 to 2014 has one less header "Power Play"
* rows that represent drawings from 2012 to 2014 have value "0" where header is "Power Play" / column position is 11
* column 1 is not useful

# add headers 
```{r}
names(raw_data) <- c("game_name", "month", "day", "year", "num1", "num2", "num3", "num4", "num5", "powerball", "power_play")
head(raw_data)
```

# list of potential questions we can ask / potential visualizations
* distribution of numbers across day, month, year
* distribution of numbers (how "random"/normally distributed is number selection?)
* what is prize size across time? (would need joining against secondary data source: https://www.txlottery.org/export/sites/lottery/Games/Powerball/Winning_Numbers/print.html_2013354932.html)
* what is powerball distribution across time? 
* what is the range of numbers within each of the 5 numbers? against powerball? against power play?
* what is likelihood of out-of-state winner?

*** don't need this - clean up later. 
#write to CSV file for upload to grading site. Will also upload sql #script used to create the DB #table.
# good site http://sphweb.bumc.bu.edu/otlt/MPH-Modules/BS/R/R-Manual/R-Manual5.html

write.csv(df, 'fltbycity.csv',row.names=FALSE)
*** don't need above - clean up later. 

# https://tibble.tidyverse.org/
# Tidy work. Use Control Shift M for #shortcut to pipes

```{r}
library(tidyverse)
library(dplyr)
```

```{r}
pb_df <- as_tibble(df)
pb_df

```

#check gather worked on each position to create a data frame for each num position
#pb_df %>% gather(num1, count, -Game_Name, -Month, -Day, -Year, -Num2,-Num3,-Num4, -Num5, - Powerball)
# select gets rid of Game Name field
#spread to widen 

```{r}

# Group the data by position - not working yet

pbnum1 <- pb_df %>% gather(Num1, count, -Game_Name, -Month, -Day, -Year, -Num2,-Num3,-Num4, -Num5, -Powerball) 
pbnum1
```
%>% select (Num1,count)


***** BELOW IS GARBAGE needs to be deleted cleaned - oly left for reference to syntax
pbnum1<- pb_df %>% select(Num1, count) %>% group_by(Num1) 
pbnum2 <- pb_df %>% gather(Num2, count, -Game_Name, -Month, -Day, -Year, -Num1,-Num3,-Num4, -Num5, -Powerball) %>% select (Num2, count)
pbnum3 <- pb_df %>% gather(Num3, count, -Game_Name, -Month, -Day, -Year, -Num1,-Num2,-Num4, -Num5, -Powerball) %>% select (Num3, count)
pbnum4 <- pb_df %>% gather(Num4, count, -Game_Name, -Month, -Day, -Year, -Num1,-Num2,-Num3, -Num5, -Powerball) %>% select (Num4, count)
pbnum5 <- pb_df %>% gather(Num5, count, -Game_Name, -Month, -Day, -Year, -Num1,-Num2,-Num3, -Num4, -Powerball) %>% select (Num5, count)
pbPball <- pb_df %>% gather(Powerball, count, -Game_Name, -Month, -Day, -Year, -Num1,-Num2,-Num3, -Num4, -Num5) %>% select (Powerball, count)

# Add Percent on time to data frame fltdata2 and total count

fltdata3 <- fltdata2 %>% mutate( percontime = on_time/(on_time + delayed), total_flights = (on_time + delayed))
fltdata3

  ggplot(fltdata3, aes(fill=airline, y=percontime, x=city)) +
    ggtitle("Ontime by City") +
    theme(plot.title = element_text(hjust = 0.5)) + 
    geom_bar(position='dodge', stat="identity") +
    xlab('City') + 
    ylab('Percent of Flights On time')
```

Conclusion:
1) Alaska airlines has less delays in each airport
2) Philly appears to be the most on time of all the airports for both airlines from a percentage perspective
3) San Francisco seems to be the most delayed airport from the graph for both airlines. 
4) I would fly Alaska if I were concerned with being at an airport on time. 